\bf Model Visualization Techniques for a Social Network Model

### 2018-07-26 paper-child.Rnw
remove some lines
proofing

Social networks, such as collaboration networks between academic researchers or friendship networks in a school or university have been studied for decades. Statistical models such as exponential random graph models (ERGM) and latent space models (LSM) are often used when studying one observation of a network when trying to model its structure. Social networks can also be observed at many points in time, and ERGM or LSM may not be appropriate for modelling the changes because they do not model network change as time passes. When studying \textit{dynamic networks}, networks observed at many time points, we need a model that allows for variation in the network over time. Models for dynamic social networks hold a great deal of potential because of how realistic they can be. Social networks do not form spontaneously, but rather they \textit{evolve} over time as edges are formed and dissolved, and new nodes join the social structure. Modeling the underlying mechanisms that create network changes in time is very complex but also provides potential to uncover hidden truths.

\par To model dynamic social networks, we use stochastic actor-oriented models (SAOMs), introduced by \citet{saompaper}. These models are different from other network models because they allow us to add node level information into the model with the traditional network structure information. Adding node-level information into the model is a more intuitive approach to model changes in a social network. For example, we expect people with common interests to be more likely to form relationships, and SAOMs allow us to incorporate this information on shared interests in the modeling process.

\par Unlike other network models, SAOMs are not very well understood. They are relatively new, especially compared to the classic ERGMs, and they are not very tractable analytically. Likelihood functions quickly become very complex objects to analyze due to the dependency structure inherent in the data. Therefore, computational methods, such as Markov chain Monte Carlo (MCMC) are used to fit models. SAOMs are typically fit to dynamic social network data using a series of MCMC phases for finding method of moments (MoM) estimates of model parameters. In order to estimate the parameters of SAOMs, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, which was developed by \citet{RSiena}. This software is a considerable contribution to the field of social network analysis, but the many simulations involved in parameter estimation are largely hidden from the software user. In order to better understand the model-fitting process of \texttt{RSiena}, we attempt to bring model fits and the fitting process to the foreground by visualizing the underlying methods and structures. Utimately, we aim to help researchers working with the models better understand the structure of and results from these models.

\par To learn more about the SAOM fitting process, we use model visualization techniques (see \citet{modelvis}) to display the model in the data space, view collections of models instead of single models, and explore the algorithms fitting the SAOMs instead of studying only the final output. SAOMs are an excellent example of a family of models that can reap many benefits from the application of model visualization techniques. For instance, the models themselves include a continuous-time Markov chain (CTMC) that is completely hidden from the analyst in the model fitting process. Visualizing the CTMC can provide researchers with more insight into the underlying features of the model. In addition, SAOMs can include a great deal of parameters in the model structure, each of which is attached to a network statistic. These statistics are often somewhat, if not highly, correlated, which causes high correlation between the corresponding parameters in a SAOM. By visualizing collections of SAOMs, we gain a better understanding of these correlations and can find ways to account for their effects in the model. Furthermore, the estimation of the parameters in a SAOM rely on convergence checks based on simulations from the fitted model. These simulations are hidden and unsaved by default for computational efficiency. With the help of both static and dynamic visualizations, we bring the hidden model fitting processes into the foreground, eventually leading to a better understanding and higher accessibility of stochastic actor-oriented models for social network analysts.

The remainder of this paper is organized as follows: In Section~\ref{sec:netintro}, we introduce basic concepts of networks and network visualizations. In Section~\ref{sec:saoms}, we formally define the family of stochastic actor-oriented models for social network analysis. In Section~\ref{sec:ModelVis}, we combine concepts from Sections~\ref{sec:netintro} and \ref{sec:saoms} in an application of the model-vis paradigm, and conclude with a discussion in Section~\ref{sec:discussion}.

@@

We provide a brief introduction to the structure of network data and common network visualization methods.

\subsubsection{Introduction to Network Structures}

Network data across fields have similar data structures. There are always units of observation and connections between those units. In a social network, the units of observations, called \textit{nodes} or \textit{actors}, may be people, while the connections, called \textit{edges} or \textit{ties}, are the relationships between people. Networks often change over time, like when new relationships are formed in a social network. The nodes and edges themselves can also have inherent variables of interest, e.g.\ the age, gender, and political affiliation of people in a social network and the type of relationship (friends, married, etc.).

These two network data structures pose unique problems to analysts, such as, ``How does the strength of a tie between two nodes affect the overall structure of the network?" or ``Do differences in node variables affect the formation of edges?" We propose to answer these questions and more through visual exploration of network data and models.

\subsubsection{Visualizing Network Data} \label{sec:netvizintro}

Network visualization, also called network mapping, is a prominent subfield of network analysis. Visualizing network data is uniquely difficult because of the structure of the data itself.  Most data visualizations rely on well-defined axes inherited from the data, such as Cartesian coordinates or spatial locations. Network data, however, carry no such inherent structure.

@@

There are two primary methods used to visualize networks that are very different from most traditional data visualizations. They are the node-link diagram and the adjacency matrix visualization \citep{knuth2000, adjmatpro}. For example, consider a network with five nodes, $\{1,2,3,4,5\}$, connected by five directed edges:  $\{2 \to 4, 3 \to 4, 1 \to 5, 3 \to 5, 5 \to 4\}$. This data is shown using the two visualization methods in Figure~\ref{fig:toyplots}.

The first method, the node-link diagram, represents nodes with points in two dimensions and then represents edges by connecting the points with lines. These lines can also have arrows on them indicating the direction of the edge for directed networks, as show in Figure~\ref{fig:toyplots}. Because there is usually no natural placement of the points (the exception is nodes with spatial locations), a random placement of the points is used, then adjusted with a layout algorithm, of which there are many \citep{drawingalgs}.

\par Some commonly used layout algorithms, such as  the Kamada-Kawai (KK) layout \citep{kamadakawai} and the Fruchterman-Reingold (FR) layout \citep{fruchterman-reingold}, are designed to mimic physical systems, drawing the graphs based on the ``forces" connecting them. Other algorithms use multi-dimensional scaling or properties of the adjacency matrix (e.g. its eigenstructure), to place the nodes \citep{drawingalgs}. In Figure~\ref{fig:toyplots}, we use the KK algorithm, and unless otherwise stated, all other node-link diagrams in this paper will use the KK layout.

\par The second method for network visualization uses the adjacency matrix of the network. The adjaceny matrix of a network, $\mathbf{A}$ is defined as follows: an entry $A_{ij}$ of $\mathbf{A}$, for two nodes $i\neq j$ in the network is defined as
\begin{equation}\label{eq:adjmat}
  A_{ij} =
  \begin{cases}
                                   1 &  \text{if an edge exists } i\to j \\
                                   0 & otherwise
  \end{cases}
\end{equation}
We only consider the presence or absence of an edge, but if the network has weighted edges, the entries in the adjacency matrix are the edge weights. An adjacency matrix visualization for our toy example is also shown in Figure~\ref{fig:toyplots}. This visualization shows the network as a grid, with each row and column representing a node. The grid space for edge $i\rightarrow j$ is filled in if $A_{ij} = 1$ and is empty if $A_{ij} = 0$.

@@

A Stochastic Actor-Oriented Model (SAOM) is a model that incorporates all three data levels of dynamic networks: actor, tie, and temporal information. It models the change of a network in time, while accounting for changes in network structure due to actor-level covariates. This model was first introduced by Snijders in 1996 \citep{saompaper}. The two titular properties of SAOMs, stochasticity and actor-orientation, are key to understanding networks as they exist naturally. Most social networks, even holding constant the set of actors over time, are ever-changing as relationships decay or grow in seemingly random ways. Actors in social networks also have inherent properties that could affect how they change their role within the network, and existing ties may affect formation or dissolution of other ties.

@@

We use the term \textit{dynamic network} to refer to a network, consisting of a fixed set of $n$ nodes, that is changing over time, and is observed at $M$ discrete time points, $t_1, \dots, t_M$ with $t_1 \le t_2 \le \dots \le t_M$. We denote the network observation at timepoint $t_k$ by $x(t_k)$. The SAOM assumes that this longitudinal network of discrete observations is embedded within a continuous time Markov chain (CTMC), which we will denote $X(T)$. This process is almost entirely unobserved: we assume that the beginning of the process, $X(0)$, is equivalent to the first network observation $x(t_1)$, while the end of the process, $X(\infty)$, is equivalent to the last observation $x(t_M)$. Nearly all other steps in the chain are unseen, with the exception of $x(t_2), \dots, x(t_{M-1})$. Unlike the first and last observations of the network, these ``in-between" observations do not have direct correspondence with steps in the continuous time Markov chain. Thus, the ``in-between" observations are considered to be ``snapshots" of the network at some point between two steps in the CTMC.

The CTMC process $X(T)$ is a series of single tie changes that happen according to a pre-defined rate function, where one actor at a time is selected and it adds or removes one outgoing tie, or does nothing. Once an actor is chosen at random according to the rate function, it is ``given" the chance to change a tie, and it tries to maximize its utility function based on the current and possible future states of the network. \\

\noindent \textbf{The Rate Function} In general, each actor $i$ in the network $x$ may change its ties, $x_{ij}$, to other nodes $j \neq i$ according to the SAOM's  \emph{rate function}, $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$, where $\boldsymbol{\alpha}$ are the parameters in the function $\rho$, and $x$ is the current network state, with covariates of interest $\mathbf{z}$. With this general formulation, it is possible that each node will have a different rate of change. We assume a simple rate function, $\rho(x, \mathbf{z}, \boldsymbol{\alpha}) = \alpha_m$ that is constant across all actors between time $t_m$ and $t_{m+1}$. With this simple rate function, we have just one set of rate parameters in the overall model. In this simple model, the rate parameters dictate how often an actor $i$ ``gets an opportunity" to change one of its ties, $x_{ij}$, for $j \in \{1, \dots, n\}$ in the time period from $t_{m}$ to $t_{m+1}$ for $m = 1, \dots, M-1$ (Note that when $j = i$, no change is made.) We also assume that the actors $i$ are conditionally independent given their ties, $x_{i1}, \dots, x_{in}$ at the current network state. Let $\tau(i|x,m) $ be the wait time until actor $i$ gets to the opportunity to change its current set of ties. For any time point, $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to the next change made by actor $i$ is exponentially distributed with expected value $\alpha_m^{-1}$. Then, we express this in Equation~\ref{eq:c1}.

@@

\noindent \textbf{The Objective Function} Because of the conditional independence assumptions given in Equation~\ref{eq:c1}, we can consider the objective function for each node separately, as only one tie from one node is changing at a time in the CTMC. The node $i$, which is the node that is given the opportunity to change at the current time point, is called the \textit{ego} node. It has the potential to interact with all other nodes in the network, $j \neq i$. These nodes $j$, are referred to as \textit{alter} nodes, and do not change any of their ties when $i$ may change. For the ego node $i$ in the current network state $x$, its \emph{objective function}, which it tries to maximize, is written as
\begin{equation} \label{eq:of1}
f_i(\boldsymbol{\beta}, x) = \sum_k \beta_k s_{ik}(x, \mathbf{Z}),
\end{equation}
for $x \in \mathcal{X}$, the space of all possible directed networks with the $n$ nodes, and $\mathbf{Z}$, the matrix of covariates. The vector $\boldsymbol{\beta}$ contains the parameters of the model with corresponding network and covariate statistics, $s_{ik}(x, \mathbf{Z})$, for $k = 1,\dots, K$. To increase the value of $f_i$, $i$ will destroy one tie where $x_{ij} = 1$, create one tie where $x_{ij}=0$, or make no change if any changes result in lower values of $f_i$.

According to \citet[p. 371]{snijders01}, there should be at least two parameters included in the model: $\beta_1$ for the outdegree of a node, and $\beta_2$ for the number of reciprocal ties held by a node. (These effects are also used in ERGMs.) The outdegree represents the propensity of nodes with a lot of outgoing ties to form more outgoing ties (the "rich get richer" effect), and the reciprocity parameter measures the tendency of outgoing ties to be returned within a network. The statistics corresponding to these two parameters, $s_{i1}(x)$ and $s_{i2}(x)$ are given in Figure~\ref{tab:effects}. In the \texttt{RSiena} software, there are over 80 possible $\boldsymbol{\beta}$ parameters to add to the model. The formulas for the corresponding statistics for all effects are provided in \citet{RSienamanual}.

The $\boldsymbol{\beta}$ parameters are either structural or covariate parameters. The structural parameters, such as the outdegree and reciprocity parameters, correspond to statistics that are only functions of the current network state, $x$. They model underlying mechanisms of network change, answering questions such as, ``How does the existing network structure influence change in the network?" The covariate parameters are functions of $x$ and additional node-level covariates of interest, $\mathbf{Z}$. A table of some possible structural and covariate parameters and their corresponding statistics is given in Table~\ref{tab:effects}.

\begin{table}
\centering
\begin{tabular}{m{4.5cm}m{4.5cm}m{3cm}}
\multicolumn{3}{l}{\textbf{Structural Effects}} \\
Name & Statistic & Figure \\
\hline\hline
outdegree  & $s_{i1}(x) = \sum_j x_{ij}$ & \includegraphics[width=2cm]{img/outdegree}\\ \hline
reciprocity  & $s_{i2}(x) = \sum_j x_{ij}x_{ji}$ & \includegraphics[width=2cm]{img/reciprocity}\\ 
%\hline
%transitive triplets  & $s_{i3}(x) = \sum_{j,h} x_{ij}x_{jh}x_{ih}$ & \includegraphics[width=1.5cm]{img/transtrip}  \\
%\\
\multicolumn{3}{l}{\textbf{Covariate Effects}}\\  
Name & Statistic & Figure \\
\hline \hline
covariate-alter  & $s_{i4}(x) = \sum_j x_{ij}z_j$ & \includegraphics[width=3cm]{img/covaralter} \\  \hline
covariate-ego  & $s_{i5}(x) = z_i\sum_j x_{ij}$ & \includegraphics[width=3cm]{img/covarego} \\ % \hline
%same covariate & $s_{i6}(x) = \sum_j x_{ij} \mathbb{I}(z_i = z_j)$ & \includegraphics[width=3cm]{img/covarsim}%\\
%jumping transitive triplets  & $s_{i7}(x) = \sum_{j \neq h} x_{ij}x_{ih}x_{hj} \mathbb{I}(z_i = z_h \neq z_j)$
\end{tabular}
\caption{\label{tab:effects} Examples of effects included in a SAOM. The dark nodes in the Figures are the ego nodes, and all others are alters. The dotted lines are ties the ego node is en(dis)couraged to make when the parameter corresponding to each effect is positive (negative). Node shapes represent levels of the covariate $z$.}
\end{table}

The objective function also defines the probability that actor $i$, when it is the ego node, chooses to change the tie to actor $j$, $p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z})$. Let $x(i \leadsto j)$ denote the network identical to $x$ with the exception of tie $x_{ij}$. It has changed to $1-x_{ij}$. The probility that the tie $x_{ij}$ changes is defined as:

\begin{equation}\label{eq:pij}
p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z}) = \dfrac{\exp\left\{f_i(\boldsymbol{\beta}, x(i\leadsto j), \mathbf{Z})\right\}}{\sum_{h \neq i} \exp\left\{f_i(\boldsymbol{\beta}, x(i \leadsto h), \mathbf{Z})\right\}}
\end{equation}
When $i = j$, the numerator represents the exponential of the value of the objective function when evaluated at the current network state. When the value of the objective function is high at the current state, the probability of not making a change is also high. In the CTMC, when actor $i$ may make a change, it chooses which tie $x_{i1, \dots, x_{in}}$ to change at random according to the probabilities $p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z})$. The objective function and the resulting values of $p_{ij}$ are combined with the rate function to fully describe the CTMC that is used to model network change in a SAOM. For more on CTMCs, see \citet{ctmc}. We explore these probabilities further in Section~\ref{sec:algorithms}

@@

\par The underlying SIENA software uses a Robbins-Monro algorithm (see \citet{robbinsmonro}) to estimate the solution of the moment equation
$$ E_{\theta}S = s_{obs}$$
where $\theta = (\boldsymbol{\alpha}, \boldsymbol{\beta})$ is the vector of rate and objective function parameters, and $s_{obs}$ is the observed vector of model statistics. The full SIENA MoM estimation algorithm operates in three phases, as described in \citet{RSienamanual, SienaAlgs}. The first phase performs initial estimation of the score functions for use in the Robbins-Monro procedure for MoM estimation. The second phase carries out the Robbins-Monro algorithm and obtains estimates of the parameter values through iterative updates and simulation from the CTMC at iterative parameter values. The third phase uses the parameter vector estimated in phase two to estimate the score functions and covariance matrix of the vector of parameter estimates and carries out convergence checks.

In each of the the first two phases of the algorithm, the procedure uses ``microsteps" that are simulated from the model as it exists in its current state in order to update either the score functions or the parameter estimates. These simulated microsteps are observed instances of the CTMC that is the backbone of the SAOM. In Section~\ref{sec:ModelVis}, we further explore these phases in the SIENA method-of-moments algorithm through visualization, bringing them out of their ``black box" and into the light.

@@

\section{Example Data and Models}

@@

Because we require binary edges for SAOMs, we focus only on very strong collaborations. For our senate collaboration networks, $x$, edges are defined as
\begin{equation}\label{eq:wpc2}
  x_{ij} =
  \begin{cases}
                                   1 &  \text{if } WPC_{ij} > 0.25 \\
                                   0 & \text{if } WPC_{ij} \leq 0.25.
  \end{cases}
\end{equation}
The node-link diagram for the four senates during the Obama administration are shown in Figure~\ref{fig:senateplot}. We can see that Senate 111 is much more densely connected than the other three, which have one large, sparsely connected component with a few smaller connected groups throughout. Some prominent names are shown to demonstrate their place in the network.

>>>>

Social networks, such as collaboration networks between academic researchers or friendship networks have been studied for decades. Statistical models such as exponential random graph models (ERGM) and latent space models (LSM) are often used when studying one observation of a network when trying to model its structure \citep{ergm, lsm}. Social networks can also be observed at many points in time, and ERGM or LSM may not be appropriate for modelling the changes because they do not model network change as time passes. When studying \textit{dynamic networks} observed at many time points, we need a model that allows for variation over time. Models for dynamic social networks hold a great deal of potential because of how realistic they can be: social networks evolve over time as edges are formed and dissolved and new nodes join. Modeling the underlying mechanisms that create network changes in time is very complex but also provides potential to uncover hidden truths.

\par One set of models for dynamic social networks, are stochastic actor-oriented models (SAOMs), introduced by \citet{saompaper}. These models are different from other dynamic network models because they allow us to add node level information into the model with the traditional network structure information. Adding node-level information into the model is a more intuitive approach to model changes in a social network because, for example we expect people with common interests to be more likely to form relationships. SAOMs allow us to incorporate this information on shared interests in the modeling process.

\par Unlike other network models, SAOMs are not very well understood. They are relatively new, especially compared to the classic ERGMs, and they are not very tractable analytically. Likelihood functions quickly become very complex due to the dependency structure inherent in the data, and so computational methods, such as Markov chain Monte Carlo (MCMC) are used to fit models. SAOMs are typically fit to dynamic social network data using a series of MCMC phases for finding method of moments (MoM) estimates of model parameters. In order to estimate the parameters of SAOMs, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena} by \citet{RSiena}. This software is a considerable contribution to the field of social network analysis, but the parameter estimation process is largely hidden from the user. In order to better understand the model-fitting process of \texttt{RSiena}, we visualize the underlying methods and structures, and utimately aim to help researchers working with SAOMs better understand the structure of and results from these models.

\par To learn more about the SAOM fitting process, we use model visualization techniques (see \citet{modelvis}) to display the model in the data space, view collections of models instead of single models, and visually explore the algorithms fitting the SAOMs. SAOMs are a model family that can reap many benefits from the application of model visualization techniques. For instance, the models themselves include a continuous-time Markov chain (CTMC) that is completely hidden from the analyst in the model fitting process. Visualizing the CTMC can provide researchers with more insight into the underlying features of the model. In addition, SAOMs can include a great deal of parameters in the model structure, each of which is attached to a network statistic. These statistics are often somewhat, if not highly, correlated, which causes high correlation between the corresponding parameters in a SAOM. By visualizing collections of SAOMs, we gain a better understanding of these correlations and can find ways to account for their effects in the model. Furthermore, the estimation of the parameters in a SAOM rely on convergence checks based on simulations from the fitted model. These simulations are hidden and unsaved by default for computational efficiency. With the help of both static and dynamic visualizations, we bring the hidden model fitting processes into the foreground, eventually leading to a better understanding and higher accessibility of stochastic actor-oriented models for social network analysts.

%The remainder of this paper is organized as follows: In Section~\ref{sec:netintro}, we introduce basic concepts of networks and network visualizations. In Section~\ref{sec:saoms}, we formally define the family of stochastic actor-oriented models for social network analysis. In Section~\ref{sec:ModelVis}, we combine concepts from Sections~\ref{sec:netintro} and \ref{sec:saoms} in an application of the model-vis paradigm, and conclude with a discussion in Section~\ref{sec:discussion}.

@@

We provide a brief introduction to the structure of network data and common network visualization methods. \\

\noindent \textbf{Introduction to Network Structures:} Network data across fields have similar data structures. There are always units of observation and connections between those units. In a social network, the units of observations, called \textit{nodes} or \textit{actors}, may be people, while the connections, called \textit{edges} or \textit{ties}, are the relationships between people. Networks often change over time, like when new relationships are formed in a social network. The nodes and edges themselves can also have inherent variables of interest, e.g.\ the age, gender, and political affiliation of people in a social network and the type of relationship (friends, married, etc.). \\

%These two network data structures pose unique problems to analysts, such as, ``How does the strength of a tie between two nodes affect the overall structure of the network?" or ``Do differences in node variables affect the formation of edges?" We propose to answer these questions and more through visual exploration of network data and models. \\

\noindent \textbf{Visualizing Network Data:} Network visualization, also called network mapping, is a prominent subfield of network analysis. Visualizing network data is uniquely difficult because of the structure of the data itself.  Most data visualizations rely on well-defined axes inherited from the data, such as Cartesian coordinates or spatial locations, but network data typically do not have an inherent structure.

@@

There are two primary methods used to visualize networks: the node-link diagram and the adjacency matrix visualization \citep{knuth2000, adjmatpro}. For example, consider a network with five nodes, $\{1,2,3,4,5\}$, connected by five directed edges:  $\{2 \to 4, 3 \to 4, 1 \to 5, 3 \to 5, 5 \to 4\}$. This data is shown using the two visualization methods in Figure~\ref{fig:toyplots}.

The first method, the node-link diagram, represents nodes with points in two dimensions and then represents edges by connecting the points with lines. These lines can also have arrows on them indicating the direction of the edge for directed networks, as show in Figure~\ref{fig:toyplots}. Because there is usually no natural placement of the points, a random placement of the points is used, then adjusted with a layout algorithm, of which there are many \citep{drawingalgs}. Some commonly used algorithms, such as the Kamada-Kawai (KK) layout \citep{kamadakawai} and the Fruchterman-Reingold (FR) layout \citep{fruchterman-reingold}, are designed to mimic physical systems, drawing the graphs based on the ``forces" connecting them. %Other algorithms use multi-dimensional scaling or properties of the adjacency matrix (e.g. its eigenstructure), to place the nodes \citep{drawingalgs}. 
In Figure~\ref{fig:toyplots}, we use the KK algorithm, and unless otherwise stated, all other node-link diagrams in this paper will use the KK layout.

\par The second method for network visualization uses the adjacency matrix of the network. The adjaceny matrix of a network, $\mathbf{A}$ is defined as follows: an entry $A_{ij}$ of $\mathbf{A}$, for two nodes $i\neq j$ in the network is defined as
\begin{equation}\label{eq:adjmat}
  A_{ij} =
  \begin{cases}
                                   1 &  \text{if an edge exists } i\to j \\
                                   0 & otherwise
  \end{cases}
\end{equation}
In this paper, we only consider the presence or absence of an edge, but if the network has weighted edges, $A_{ij}$ is the weight of the edge from $i \to j$. An adjacency matrix visualization for our toy example is also shown in Figure~\ref{fig:toyplots}. This visualization shows the network as a grid, with each row and column representing a node. The grid space for edge $i\rightarrow j$ is filled in if $A_{ij} = 1$ and is empty if $A_{ij} = 0$.

@@

A Stochastic Actor-Oriented Model (SAOM) is a model that incorporates all three data levels of dynamic networks: actor, tie, and temporal information. It models the change of a network in time, while accounting for changes in network structure due to actor-level covariates. The two titular properties of SAOMs, stochasticity and actor-orientation, are key to understanding networks as they exist naturally. Most social networks, even holding constant the set of actors over time, are ever-changing as relationships decay or grow in seemingly random ways. Actors in social networks also have inherent properties that could affect how they change their role within the network, and existing ties may affect formation or dissolution of other ties.

@@

We use the term \textit{dynamic network} to refer to a network, consisting of a fixed set of $n$ nodes, that is changing over time, and is observed at $M$ discrete time points, $t_1, \dots, t_M$ with $t_1 < t_2 < \dots < t_M$. We denote the network observation at timepoint $t_m$ by $x(t_m)$ for $m = 1, \dots M$. The SAOM assumes that this longitudinal network of discrete observations is embedded within a CTMC, which we will denote by $X(T)$. This CTMC is almost entirely unobserved: we assume that the beginning of the process, $X(0)$, is equivalent to the first network observation $x(t_1)$, while the end of the process, $X(\infty)$, is equivalent to the last observation $x(t_M)$. Nearly all other steps in the chain are unseen, with the exception of $x(t_2), \dots, x(t_{M-1})$. Unlike the first and last observations of the network, these ``in-between" observations do not have direct correspondence with steps in the continuous time Markov chain. Thus, the observations $x(t_2), \dots, x(t_{M-1})$ are considered to be ``snapshots" of the network at some point between two steps in the CTMC.

The CTMC process $X(T)$ is a series of single tie changes that happen according to the SAOM's pre-defined rate function, where one actor at a time is selected and it adds or removes one outgoing tie, or does nothing. Once an actor is chosen at random according to the rate function, it is ``given" the chance to change a tie, and it tries to maximize its utility function based on the current and currently reachable states of the network. \\

\noindent \textbf{The Rate Function} In general, each actor $i$ in the network $x$ gets an opportunity to changes its ties, $x_{ij}$, to other nodes $j \neq i$ according to the SAOM's \emph{rate function}, $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$, where $\boldsymbol{\alpha}$ are the parameters in the function $\rho$, and $x$ is the current network state, with covariates of interest $\mathbf{z}$. With this general formulation, it is possible that each node will have a different rate of change. We assume a simple rate function, $\rho(x, \mathbf{z}, \boldsymbol{\alpha}) = \alpha_m$ that is constant across all actors between time $t_m$ and $t_{m+1}$. With this simple rate function, we have just one set of rate parameters in the overall model. In this simple model, the rate parameters dictate how often an actor $i$ ``gets an opportunity" to change one of its ties, $x_{ij}$, for $j \in \{1, \dots, n\}$ in the time period from $t_{m}$ to $t_{m+1}$ for $m = 1, \dots, M-1$ (Note that if $j = i$, no change is made.) A SAOM also assumes that the actors $i$ are conditionally independent given their ties, $x_{i1}, \dots, x_{in}$ at the current network state. Let $\tau(i|x,m) $ be the wait time until actor $i$ gets to the opportunity to change its current set of ties. For any time point, $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to the next change made by actor $i$ is exponentially distributed with expected value $\alpha_m^{-1}$, as shown in Equation~\ref{eq:c1}.

@@

\noindent \textbf{The Objective Function} Because of the conditional independence assumptions given in Equation~\ref{eq:c1}, we can consider the objective function for each node separately, as only one tie from one node is changing at a time in the CTMC. The node $i$, which is given the opportunity to change at the current time point, is called the \textit{ego} node. It has the potential to interact with all other nodes in the network, $j \neq i$. These nodes $j$, are referred to as \textit{alter} nodes, and do not change any of their ties when $i$ may change. For the ego node $i$ in the current network state $x$, its \emph{objective function}, which it tries to maximize, is written as
\begin{equation} \label{eq:of1}
f_i(\boldsymbol{\beta}, x, \mathbf{Z}) = \sum_{k=1}^K \beta_k s_{ik}(x, \mathbf{Z}),
\end{equation}
for $x \in \mathcal{X}$, the space of all possible directed networks with the $n$ nodes, and $\mathbf{Z}$, the matrix of covariates, and $K$ the total number of parameters in the objective function. The vector $\boldsymbol{\beta}$ contains the parameters of the model with corresponding statistics, $s_{ik}(x, \mathbf{Z})$, for $k = 1,\dots, K$. To increase the value of $f_i$, $i$ will destroy one tie where $x_{ij} = 1$, create one tie where $x_{ij}=0$, or make no change if all changes will result in lower values of $f_i(\boldsymbol{\beta}, x, \mathbf{Z})$.

According to \citet[p. 371]{snijders01}, there should be at least two parameters included in the model: $\beta_1$ for the outdegree of nodes, and $\beta_2$ for the reciprocity of nodes. (These effects are also used in ERGMs.) The outdegree represents the propensity of nodes with a lot of outgoing ties to form more outgoing ties (the ``rich get richer" effect), and the reciprocity parameter measures the tendency of outgoing ties to be returned within a network. The statistics corresponding to these two parameters, $s_{i1}(x)$ and $s_{i2}(x)$ are given inTable~\ref{tab:effects}. In the \texttt{RSiena} software, there are over 80 possible $\boldsymbol{\beta}$ parameters to add to the model. The formulas for the corresponding statistics for all effects are provided in \citet{RSienamanual}.

The parameters of $f_i(\boldsymbol{\beta}, x, \mathbf{Z})$ are either structural or covariate parameters. The structural parameters, such as the outdegree and reciprocity parameters, correspond to statistics that are \emph{only} functions of the current network state, $x$. They model underlying mechanisms of network change, answering questions such as, ``How does the existing network structure influence change in the network?" The covariate parameters are functions of $x$ and additional node-level covariates of interest, $\mathbf{Z}$. Some possible covariate parameters and their corresponding statistics are discussed in Section~\ref{sec:exdatamods} and shown in Figure~\ref{fig:structures.}

\begin{table}
\centering
\begin{tabular}{m{4.5cm}m{4.5cm}m{3cm}}
\multicolumn{3}{l}{\textbf{Structural Effects}} \\
Name & Statistic & Figure \\
\hline\hline
outdegree  & $s_{i1}(x) = \sum_j x_{ij}$ & \includegraphics[width=2cm]{img/outdegree}\\ \hline
reciprocity  & $s_{i2}(x) = \sum_j x_{ij}x_{ji}$ & \includegraphics[width=2cm]{img/reciprocity}\\ 
%\hline
%transitive triplets  & $s_{i3}(x) = \sum_{j,h} x_{ij}x_{jh}x_{ih}$ & \includegraphics[width=1.5cm]{img/transtrip}  \\
%\\
% \multicolumn{3}{l}{\textbf{Covariate Effects}}\\  
% Name & Statistic & Figure \\
% \hline \hline
% covariate-alter  & $s_{i4}(x) = \sum_j x_{ij}z_j$ & \includegraphics[width=3cm]{img/covaralter} \\  \hline
% covariate-ego  & $s_{i5}(x) = z_i\sum_j x_{ij}$ & \includegraphics[width=3cm]{img/covarego} \\ % \hline
%same covariate & $s_{i6}(x) = \sum_j x_{ij} \mathbb{I}(z_i = z_j)$ & \includegraphics[width=3cm]{img/covarsim}%\\
%jumping transitive triplets  & $s_{i7}(x) = \sum_{j \neq h} x_{ij}x_{ih}x_{hj} \mathbb{I}(z_i = z_h \neq z_j)$
\end{tabular}
\caption{\label{tab:effects} Examples of structural effects included in a SAOM. The dark nodes in the Figures are the ego nodes, and all others are alters. The dotted lines are ties the ego node is en(dis)couraged to make when the parameter corresponding to each effect is positive (negative).}
\end{table}

The objective function also defines the probability that actor $i$, when it is the ego node, chooses to change the tie to actor $j$, denoted by $p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z})$. Let $x(i \leadsto j)$ be the network identical to $x$ with the exception of tie $x_{ij}$, which is equal to $1-x_{ij}$. The probility that the tie $x_{ij}$ changes is defined as:

\begin{equation}\label{eq:pij}
p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z}) = \dfrac{\exp\left\{f_i(\boldsymbol{\beta}, x(i\leadsto j), \mathbf{Z})\right\}}{\sum_{h \neq i} \exp\left\{f_i(\boldsymbol{\beta}, x(i \leadsto h), \mathbf{Z})\right\}}
\end{equation}%When $i = j$, the numerator represents the exponential of the value of the objective function when evaluated at the current network state. 
When the value of the objective function is high at the current state, the probability of not making a change is also high. In the CTMC, when actor $i$ may make a change, it chooses which tie $x_{i1}, \dots, x_{in}$ to change at random according to the probabilities $p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z})$. The objective function and the resulting values of $p_{ij}$ are combined with the rate function to fully describe the CTMC that is used to model network change in a SAOM. For more on CTMCs, see \citet{ctmc}. We explore these probabilities further in Section~\ref{sec:algorithms}

@@

\par The underlying SIENA software uses a Robbins-Monro algorithm (see \citet{robbinsmonro}) to estimate the solution of the moment equation
\begin{equation}\label{eq:mom}
E_{\theta}S = s_{obs}
\end{equation}
where $\theta = (\boldsymbol{\alpha}, \boldsymbol{\beta})$ is the vector of rate and objective function parameters, and $s_{obs}$ is the observed vector of the model statistics over all network observations. The full SIENA MoM estimation algorithm operates in three phases, as described in \citet{RSienamanual, SienaAlgs}. The first phase performs initial estimation of the score functions for use in the Robbins-Monro procedure for MoM estimation. The second phase carries out the Robbins-Monro algorithm and obtains estimates of the parameter values through iterative updates and simulation from the CTMC at iterative parameter values. The third phase uses the parameter vector estimated in phase two to estimate the score functions and covariance matrix of the vector of parameter estimates and carries out convergence checks.

In each of the the first two phases of the algorithm, the procedure uses ``microsteps" that are simulated from the model as it exists in its current state in order to update either the score functions or the parameter estimates. These simulated microsteps are realized steps in the CTMC that is the backbone of the SAOM. In Section~\ref{sec:algorithms}, we further explore these phases in the SIENA method-of-moments algorithm through visualization, bringing them out of their ``black box" and into the light.

@@

\section{Example Data and Models}\label{sec:exdatamods}

@@

Because we require binary edges for SAOMs, we focus only on very strong collaborations, where $WPC$ is high. For a senate collaboration network $x$, edges are defined as
\begin{equation}\label{eq:wpc2}
  x_{ij} =
  \begin{cases}
                                   1 &  \text{if } WPC_{ij} > 0.25 \\
                                   0 & \text{if } WPC_{ij} \leq 0.25.
  \end{cases}
\end{equation}
The node-link diagram for the four senates during the Obama administration are shown in Figure~\ref{fig:senateplot}. We can see that Senate 111 is much more densely connected than the other three, which have one large, sparsely connected component with a few smaller connected groups throughout. Some prominent names are shown to demonstrate their place in the network.
